Neon: Nuclear Norm to Beat Muon

### 2025-04-27 article/muon.tex
Updated figures

\section{Prototyping phase report}
\begin{enumerate}
    \item Update rule is derived, see idea
    \item Update rule methods are tested: power iteration vs Lanczos (see playground)
    \item Recorded the distribution of singular values of gradients during NanoGPT training (see Firgures ~\ref{fig:svd_all}, \ref{fig:svd_all}).
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/svd_evolution_all.png}}
        \caption{Singular values of random 100 parameters via 200 iterations}
        \label{fig:svd_all}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/svd_evolution.png}}
        \caption{Singular values of random 100 parameters for 5-th,45-th, 65-th,175-th and 200-th iteration}
        \label{fig:svd}
    \end{figure}

    \item NanoGPT is tested on Muon and Adam (add fig). The results are strange, because the methods do not converge at the expected rate (Muon is slower). Besides, Neon (rank-1 version) does not converge (see Figures ~\ref{fig:train_loss} \ref{fig:val_loss})
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/loss_train.png}}
        \caption{Train loss}
        \label{fig:train_loss}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/loss_val.png}}
        \caption{Validation loss}
        \label{fig:val_loss}
    \end{figure}

@@

    \item Neon (rank-1 version), Muon, AdamW and SGD are compared on MLP and CNN (see Figures~\ref{fig:mlp_epochs}, \ref{fig:cnn_epochs}, \ref{fig:mlp_time}, and~\ref{fig:cnn_time}). All methods work correctly, but again there is the problem with which one is the fastest (for now, it's SGD).
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/accuracy_vs_epochs_simple.png}}
        \caption{MLP: self.linear1 = nn.Linear(32*32*3, 512), self.linear2 = nn.Linear(512, 10), self.activ = nn.GELU()}
        \label{fig:mlp_epochs}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/accuracy_vs_epochs_moderate.png}}
        \caption{CNN: 2 convolutional blocks, 2 fully connected layers, activation + dropout}
        \label{fig:cnn_epochs}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/accuracy_vs_time_simple.png}}
        \caption{MLP: wallclock time measurements}
        \label{fig:mlp_time}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{../figs/mlp24/accuracy_vs_time_moderate.png}}
        \caption{CNN: wallclock time measurements}
        \label{fig:cnn_time}
    \end{figure}
\end{enumerate}\\

>>>>

\section{Prototyping phase report}
\begin{enumerate}
    \item Update rule is derived, see idea
    \item Update rule methods are tested: power iteration vs Lanczos (see playground)
    \item Recorded the distribution of singular values of gradients during NanoGPT training (see Firgures ~\ref{fig:svd_all}, \ref{fig:svd_all}).
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/svd_evolution_all.png}}
        \caption{Singular values of $50257\times 1280$ layer via 200 iterations}
        \label{fig:svd_all}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/svd_evolution.png}}
        \caption{Singular values of  of $50257\times 1280$ layer for 5-th,45-th, 65-th,175-th and 200-th iteration}
        \label{fig:svd}
    \end{figure}

    \item NanoGPT is tested on Muon and Adam (add fig). The results are strange, because the methods do not converge at the expected rate (Muon is slower). Besides, Neon (rank-1 version) does not converge (see Figures ~\ref{fig:train_loss} \ref{fig:val_loss})
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/loss_train.png}}
        \caption{Train loss}
        \label{fig:train_loss}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/loss_val.png}}
        \caption{Validation loss}
        \label{fig:val_loss}
    \end{figure}

@@

    \item Neon (rank-1 version), Muon, AdamW and SGD are compared on MLP and CNN (see Figures~\ref{fig:mlp_epochs}, \ref{fig:cnn_epochs}, \ref{fig:mlp_time}, and~\ref{fig:cnn_time}). All methods work correctly, but again there is the problem with which one is the fastest (for now, it's SGD).
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/accuracy_vs_epochs_simple.png}}
        \caption{MLP: self.linear1 = nn.Linear(32*32*3, 512), self.linear2 = nn.Linear(512, 10), self.activ = nn.GELU()}
        \label{fig:mlp_epochs}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/accuracy_vs_epochs_moderate.png}}
        \caption{CNN: 2 convolutional blocks, 2 fully connected layers, activation + dropout}
        \label{fig:cnn_epochs}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/accuracy_vs_time_simple.png}}
        \caption{MLP: wallclock time measurements}
        \label{fig:mlp_time}
    \end{figure}
    \begin{figure}[h!]
        \center{\includegraphics[width=0.8\linewidth]{figs/mlp24/accuracy_vs_time_moderate.png}}
        \caption{CNN: wallclock time measurements}
        \label{fig:cnn_time}
    \end{figure}
\end{enumerate}\\
